---
title: "Tidy Data Wrangling - Part A"
subtitle: "Applications of Data Science"
author: "Giora Simchoni"
institute: "Stat. and OR Department, TAU"
date: "`r Sys.Date()`"
output_dir: "images"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    seal: false
    chakra: "../libs/remark-latest.min.js"
    includes:
      in_header: "../header.html"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: logo-slide

---

class: title-slide

## Tidy Data Wrangling - Part A

### Applications of Data Science - Class 2

### Giora Simchoni

#### `gsimchoni@gmail.com and add #dsapps in subject`

### Stat. and OR Department, TAU
### `r Sys.Date()`

---
```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

class: section-slide

# `dplyr`: Basic Data Verbs

---

# Basic Data Verbs

- `filter()` rows based on one or more conditions
- `mutate()` one or more columns, usually based on existing columns
- `select()` the column(s) you want
- `arrange()` rows by one or more columns order
- `summarize()` or `summarise()` that single quantity off a column
- `pull()` a column as a vector, don't want it as a column no more

And the much beloved `group_by()`: do *whatever* by groups of one or more variables.

---
## Read in the data

```{r Basic1, message=FALSE}
library(tidyverse)

okcupid <- read_csv("../data/okcupid.csv.zip")
```

.font80percent[Reminder:]

```{r Basics2}
dim(okcupid)
colnames(okcupid)
```

---

## `mutate()`

Add a column `height_cm`, the `height` in centimeters:

```{r Basic3}
okcupid <- okcupid %>%
  mutate(height_cm = 2.54 * height)
```

.insight[
`r emo::ji("bulb")` if you also load the `magrittr` package you could do:

`okcupid %<>% mutate(height_cm = 2.54 * height)`
]

---

## `filter()` and `select()`

Filter only women, select only age and height:

```{r Basics4}
okcupid %>%
  filter(sex == "f") %>%
  select(age, height)
```

---

Same but income over 100K, and select all essay questions:

```{r Basic5}
okcupid %>%
  filter(sex == "f", income > 100000) %>%
  select(starts_with("essay"))
```

---

Same but using a range of columns:

```{r Basic5B}
okcupid %>%
  filter(sex == "f", income > 100000) %>%
  select(essay0:essay9)
```

.insight[
`r emo::ji("bulb")` Many, many such gifts, see [`tidyselect`](https://tidyselect.r-lib.org/reference/select_helpers.html)
]

---

## `summarize()`

Find the average height of women

```{r Basic6}
okcupid %>%
  filter(sex == "f") %>%
  summarize(avg_height = mean(height_cm, na.rm = TRUE))
```

Notice we got a `tibble`. We could either `pull` this single number:

```{r Basics7}
okcupid %>%
  filter(sex == "f") %>%
  summarize(avg_height = mean(height_cm, na.rm = TRUE)) %>%
  pull()
```

---

Or pull the vector of heights first, then calculate their mean:

```{r Basics8}
okcupid %>%
  filter(sex == "f") %>%
  pull(height_cm) %>%
  mean(na.rm = TRUE)
```

Amazingly, this would also work:

```{r Basic8-non-functional}
mean(pull(filter(okcupid, sex == "f"), height_cm), na.rm = TRUE)
```

---

## `group_by()`

But why settle for women only?

```{r Basics9}
okcupid %>%
  group_by(sex) %>%
  summarize(avg_height = mean(height_cm, na.rm = TRUE))
```

And you might want to consider `rename()`ing sex!

```{r Basics9B, eval=FALSE}
okcupid %>%
  group_by(sex) %>%
  summarize(avg_height = mean(height_cm, na.rm = TRUE)) %>%
  rename(gender = sex)
```

---

Group by multiple variables, get more summaries, arrange by descending average height:

```{r Basics10}
okcupid %>%
  group_by(sex, status) %>%
  summarize(avg_height = mean(height_cm, na.rm = TRUE),
            med_height = median(height_cm, na.rm = TRUE),
            n = n()) %>%
  arrange(-med_height)
```

---

## Pro tip: `count()`

When all you want is, well, `count`, no need to `group_by`:

```{r Basics-Count}
okcupid %>% count(body_type, sort = TRUE)
```

---

## Pro tip: `add_count()`

Add count without first creating an initial table, joining etc.:

```{r Basics-Add-Count}
okcupid %>%
  mutate(id = row_number()) %>%
  select(id, body_type, sex) %>%
  add_count(body_type, name = "n_bt") %>%
  filter(n_bt > 10000) %>%
  head(5)
```

---

class: section-slide

# Beyond Basics

---

## A simple answer to the religion question?

```{r Advanced1}
okcupid %>% count(religion)
```

---

## Recoding with `case_when()`

```{r Case-When}
okcupid <- okcupid %>% mutate(religion2 = case_when(
  str_detect(religion, "agnosticism") | str_detect(religion, "atheism") ~ "atheist",
  str_detect(religion, "buddhism") ~ "buddhist",
  str_detect(religion, "christianity") | str_detect(religion, "catholicism") ~ "christian",
  str_detect(religion, "judaism") ~ "jewish",
  str_detect(religion, "hinduism") ~ "hindu",
  str_detect(religion, "islam") ~ "muslim",
  TRUE ~ "NA"))

okcupid %>% count(religion2, sort = TRUE)
```

---

### Getting extreme observations with `top_n()` and `top_frac()`

```{r Top-N}
okcupid %>%
  select(sex, age) %>%
  group_by(sex) %>%
  top_n(3, age)
```

---

Unfortunately `top_n` does not deal with ties, could use `slice` for that:

```{r Top-N2}
okcupid %>%
  select(sex, age) %>%
  group_by(sex) %>%
  top_n(3, age) %>%
  slice(1:3)
```

.insight[
`r emo::ji("bulb")` Or use `rank()`
]

---

### Remove duplicates with `distinct()`

```{r Distinct}
okcupid %>%
  filter(diet == "kosher") %>%
  distinct(body_type, drugs)
```

.insight[
`r emo::ji("bulb")` `distinct()` is much more powerful than `unique()`, see `?distinct`.

To count number of distinct obs look at `n_distinct()`
]

---

### The `_at()`, `_if()` and `_all()` families

Many of the verbs we've seen come with these suffixes:

```{r Select-If}
okcupid %>%
  select_if(is.numeric)
```

Do you see something strange?

---

Take care of those missing observations for me without breaking the pipe:

```{r Select-If2}
okcupid %>%
  na_if(-1) %>%
  select_if(is.numeric)
```

---

Transform all my numeric columns with `log`:

```{r Mutate-All}
okcupid %>%
  na_if(-1) %>%
  select_if(is.numeric) %>%
  mutate_all(log)
```

.insight[
`r emo::ji("bulb")` Also see `mutate_if()`
]

---

Same but add `sqrt` and keep original columns:

```{r Mutate-All2}
okcupid %>%
  na_if(-1) %>%
  select_if(is.numeric) %>%
  mutate_all(list(logged = log, sqrted = sqrt))
```

---

Same but take care of zeros under `log`:

```{r Mutate-All3}
okcupid %>%
  na_if(-1) %>%
  select_if(is.numeric) %>%
  mutate_all(list(logged = function(x) log(x + 1), sqrted = sqrt))
```

---

Same but select only non-negative columns:

```{r Mutate-All4}
is_non_negative <- function(x) is.numeric(x) && (is.na(x) || x >= 0)

okcupid %>%
  na_if(-1) %>%
  select_if(is_non_negative) %>%
  mutate_all(list(logged = function(x) log(x + 1), sqrted = sqrt))
```

---

On second thought `log` would probably be appropriate just for `income` and `height_cm` (not really, just for demo):

```{r Mutate-At}
okcupid %>%
  na_if(-1) %>%
  mutate_at(c("income", "height_cm"),
            list(logged = function(x) log(x + 1), sqrted = sqrt)) %>%
  select(ends_with("logged"), ends_with("sqrted"))
```

---

## Dealing with `NA`s

You've already seen `na_if()`. We could simply, always, keep those `NA`s in income:

```{r NA1, eval=FALSE}
okcupid <- okcupid %>%
  mutate(income = ifelse(income == -1, NA, income))
```

Or:

```{r NA2}
okcupid <- okcupid %>%
  mutate(income = na_if(income, -1))
```

Dropping `NA`s with, well, `drop_na()`:

```{r NA3}
okcupid_no_nas <- okcupid %>% drop_na()
```

---

Replacing `NA`s with, well, `replace_na()`:

```{r NA4}
okcupid_back_to_minus1 <- okcupid %>% replace_na(list(income = -1))
```

Could be useful for imputing `NA`s, say the median:

```{r NA5}
okcupid_na_income_imputed <- okcupid %>%
  replace_na(list(income = median(.$income, na.rm = TRUE)))
```

---

### Sampling with `sample_n()` and `sample_frac()`

```{r Sample1}
okcupid %>% select(drugs, age, income, sex) %>%
  group_by(drugs) %>%
  sample_n(3, replace = TRUE)
```

---
class: section-slide

# Put it in a function

---

### Compose a function which would accept an unquoted variable

```{r Functionize1}
count_var_for_gender <- function(var, gender) {
  okcupid %>%
    filter(sex == gender) %>%
    count({{var}}, sort = TRUE)
}

count_var_for_gender(body_type, "f") %>% head(9)
```

---
### Making a `data.frame` function pipeable

```{r Functionize2}
transform_all_my_numerics <- function(df, transformation) {
  df %>% mutate_if(is.numeric, transformation)
}

okcupid %>%
  transform_all_my_numerics(log) %>%
  select_if(is.numeric)
```

---

### `invisible()`

If your function does not return a `data.frame` make it!

```{r Functionize3}
print_n_rows <- function(df) {
  cat("number of rows: ", nrow(df), "\n")
  invisible(df)
}

okcupid %>%
  filter(sex == "m", body_type %in% c("fit", "thin", "skinny")) %>%
  print_n_rows() %>%
  summarise(mean_height = mean(height_cm, trim = 0.025))
```

---

Or even better:

```{r Functionize5}
filter_and_print <- function(df, ...) {
  df_filtered <- df %>% filter(...)
  cat("number of rows: ", nrow(df_filtered), "\n")
  df_filtered
}

okcupid %>%
  filter_and_print(sex == "m", body_type %in% c("fit", "thin", "skinny")) %>%
  summarise(mean_height = mean(height_cm, trim = 0.025))
```


.insight[
`r emo::ji("bulb")` for better living see `glue::glue("number of rows: {nrow(df)}")`
]
