---
title: "Introduction to Networks"
subtitle: "Applications of Data Science"
author: "Giora Simchoni"
institute: "Stat. and OR Department, TAU"
date: "`r Sys.Date()`"
output_dir: "images"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    seal: false
    chakra: "../libs/remark-latest.min.js"
    includes:
      in_header: "../header.html"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: logo-slide

---

class: title-slide

## Introduction to Networks

### Applications of Data Science - Class 6

### Giora Simchoni

#### `gsimchoni@gmail.com and add #dsapps in subject`

### Stat. and OR Department, TAU
### `r Sys.Date()`

---
```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

class: section-slide

# Why Networks?

---

## Because this

<img src = "images/israeli_songs_excel.png" style = "width: 100%">

Can only get you so far.
---

## Divided we sing

<img src = "images/israeli_coops.png" style = "width: 100%">

---

class: section-slide

# Networks Overview

---

## A Network

A network, is comprised of:

- Nodes (vertices, points), joined together in pairs by
- Edges (links, connections)

Many types of networks:

- Physical networks: telephone lines, roads, airline routes, rivers
- Information networks: WWW, citation networks
- Social networks: Facebook, Twitter, but not just: this class, Marriage between Royal Houses
- Biological networks: "food webs" (what eats what?), metabolical networks

---

## Physical Networks

<img src = "images/gush_dan_train_system.jpg" style = "width: 60%">

.font80percent[
[מערכת להסעת המונים במטרופולין תל אביב](https://he.wikipedia.org/wiki/%D7%9E%D7%A2%D7%A8%D7%9B%D7%AA_%D7%9C%D7%94%D7%A1%D7%A2%D7%AA_%D7%94%D7%9E%D7%95%D7%A0%D7%99%D7%9D_%D7%91%D7%9E%D7%98%D7%A8%D7%95%D7%A4%D7%95%D7%9C%D7%99%D7%9F_%D7%AA%D7%9C_%D7%90%D7%91%D7%99%D7%91)
]
---

## Information Networks

<img src = "images/wikipedia_site_map.png" style = "width: 100%">

.font80percent[
[Wikipedia Main Page Site Map](https://en.wikipedia.org/wiki/Site_map)
]
---

## Social Networks

<img src = "images/the_marker_judges.png" style = "width: 80%">

.font80percent[
[חשיפה: הרשימה שכל עורך דין חייב להכיר](https://www.themarker.com/law/1.3065986)
]
---

## Biological Networks

<img src = "images/worm_neural_network.png" style = "width: 80%">

.font80percent[
[The structure of the nervous system of the nematode Caenorhabditis elegans](https://royalsocietypublishing.org/doi/abs/10.1098/rstb.1986.0056)
]
---

## Hard to classify Networks

<img src = "images/john_lennon_timeline.png" style = "width: 100%">

---

## Networks Properties

- Connected/Not:
- Directed/Not:
- Acyclic/Not:
- Self-loop-edges/Not:
- Weighted edges/Not:
- Time-varied/Not:

---

## Bipartite Networks

<img src = "images/marvel_co_occurrence.png" style = "width: 80%">

---

## Bipartite Networks

<img src = "images/marvel_bipartite_network.png" style = "width: 80%">

.font80percent[
[Adapted from: Which Marvel Characters and Movies are the Most Central? / Félix Luginbühl](https://felixluginbuhl.com/network/)
]
---

## Multilayer Networks

---

## Trees and Hierarchies

---

class: section-slide

# Intro. to Networks Math
