---
title: 'Network Models'
subtitle: 'Applications of Data Science'
author: 'Giora Simchoni'
institute: 'Stat. and OR Department, TAU'
date: '`r Sys.Date()`'
output_dir: 'images'
output:
  xaringan::moon_reader:
    css: '../slides.css'
    seal: false
    chakra: '../libs/remark-latest.min.js'
    includes:
      in_header: '../header.html'
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: logo-slide

---

class: title-slide

## Network Models

### Applications of Data Science - Class 10

### Giora Simchoni

#### `gsimchoni@gmail.com and add #dsapps in subject`

### Stat. and OR Department, TAU
### `r Sys.Date()`

---
```{r child = '../setup.Rmd'}
```

```{python, echo=FALSE}
import pandas as pd
import networkx as nx
import matplotlib.pyplot as plt
import numpy as np

def degree_hist(G, xlabel='Degree', ylabel='Freq'):
  deg = [deg for node, deg in G.degree()]
  plt.hist(deg)
  plt.xlabel(xlabel, fontsize=16)
  plt.ylabel(ylabel, fontsize=16)
```

class: section-slide

# Motivation

---

### *Degree histograms* of 3 real networks:

- The cast of RuPaul's Drag Race Twitterverse
- The Israeli artists musical cooperations in the 21st century
- The Sci-Fi themed correlation network

```{python Degree-Hists, echo=FALSE, fig.asp=0.5, out.width="80%"}
rupaul = pd.read_csv('../data/queens_edges.csv')
Ru = nx.from_pandas_edgelist(rupaul, source='from', target='to', create_using=nx.Graph)
scifi_books = pd.read_csv('../data/sci_fi_final_edgelist.csv')
Sc = nx.from_pandas_edgelist(scifi_books, 'book', 'book2', ['corr'])
israeli_artists = pd.read_csv('../data/israeli_artists_coops.csv')
I = nx.from_pandas_edgelist(israeli_artists, 'from', 'to')


plt.gcf().subplots_adjust(bottom=0.15)
plt.subplot(1, 3, 1)
degree_hist(I, xlabel='')
plt.subplot(1, 3, 2)
degree_hist(Sc, ylabel='')
plt.subplot(1, 3, 3)
degree_hist(Ru, xlabel='', ylabel='')
plt.show()
```

.insight[
`r emo::ji("bulb")` Which is which?
]

---

### Why model networks?

- Know your network better:
  - how it was formed
  - what class it belongs to (clustering)
  - how and why it deviates from model
- Predict behavior in network:
  - epidemic spread/resistance
  - search
  - link prediction
  - node disambiguation
- Generalization
- Simulations and the ability to estimate metrics on huge networks
  
---

class: section-slide

# The Erdős–Rényi model .font80percent[(Random Graphs)]

---

## You know the ER model

In a $n$ nodes undirected network, suppose edges form identically and independently from each other.

Each possible edge is a Bernoulli trial with probability $p$.

Then the no. of edges $M$ is a Binomial RV $\sim Binom({n\choose 2},p)$

.insight[
`r emo::ji("bulb")` So what is $P(M=m)$? What is $E(M)$? What is $Var(M)$?
]

The $G(n, p)$ collection of all simple networks formed this way is the Erdős–Rényi model.

---

## Mean Degree

The mean number of edges is: $E(M)={n\choose 2}p$

We have defined the mean degree to be: $c = \frac{2m}{n}$, but now we treat $m$ as a RV, so the mean degree would be:

$c=E(D)=E(\frac{2M}{n})=\frac{2E(M)}{n}=\frac{2}{n}{n\choose 2}p=(n-1)p$

.insight[
`r emo::ji("bulb")` Say it in words, it makes sense.
]

But this implies we could have formulated the ER model slightly different:

Let $D$ be the degree of a particular node in an undirected network with $n$ nodes. Suppose each node "chooses" its neighbors with probability $p$, then:  $D\sim Binom(n-1,p)$

---

Estimating $n$, $p$ and $c$ from our networks:

```{python}
def estimate_p(m, n): return m / (n * (n - 1) / 2)
def estimate_c(G): return np.mean(list(dict(G.degree()).values()))

n_isr, n_sci, n_ru = I.number_of_nodes(), Sc.number_of_nodes(), Ru.number_of_nodes()
m_isr, m_sci, m_ru = I.number_of_edges(), Sc.number_of_edges(), Ru.number_of_edges()
p_isr, p_sci, p_ru = estimate_p(m_isr, n_isr), estimate_p(m_sci, n_sci), estimate_p(m_ru, n_ru)
c_isr, c_sci, c_ru = estimate_c(I), estimate_c(Sc), estimate_c(Ru)

pd.DataFrame({
  'network': ['Israeli Artists', 'Sci-Fi Books', 'RuPaul Verse'],
  'n': [n_isr, n_sci, n_ru],
  'm': [m_isr, m_sci, m_ru],
  'p': [p_isr, p_sci, p_ru],
  'c': [c_isr, c_sci, c_ru]
})
```


---

## Degree Distribution

And since we're dealing with large *sparse* networks, we might mention that as $n\to\infty$ and $p$ is small, we expect the Binomial distribution to be approximated by the Poisson distribution, such that:

$D \sim Pois((n-1)p)$ or $D \sim Pois(c)$

Which is why the ER model is sometimes referred to the Poisson random graph.

---

## ER model In NetworkX

```{python ER, out.width="50%"}
G = nx.erdos_renyi_graph(n = 100, p = 0.3)

plt.figure()
nx.draw_networkx(G)
plt.show()
```

---

```{python ER-Deg_Dist, out.width="50%"}
deg = [deg for node, deg in G.degree()]
h = plt.hist(deg)
plt.show()
```

.insight[
`r emo::ji("bulb")` Does this degree distribution resemble any of the distributions we've seen?
]

---

## Transitivity

We have defined the Clustering Coefficient or Transitivity to be:

$Transitivity(G)=\frac{\text{#closed triads}}{\text{#triads}}$

In other words it is the probability of two neighbors of the same node to also be connected.

.insight[
`r emo::ji("bulb")` What is that probability in the ER model?

What do you think of it after what you have seen?
]

---

## Diameter

We have defined the diameter as the maximal shortest distance between any pair of nodes.

It turns out this has a nice expression in the ER model:

- Take a random node in a ER network, how many neighbors do you expect it to have?
- Take each one of these neighbors, and take each and one's neighbors - what do you get?
- Do this $l$ times, what do you get? What would you get "in the end"?

The average path length from a node to all other nodes is $\sim \frac{\ln(n)}{\ln(c)}$.

Similarly one can show the diameter is $\sim 2\frac{\ln(n)}{\ln(c)}$

---

```{python}
d_isr = nx.diameter(I)
d_sci = nx.diameter(Sc)
d_ru = nx.diameter(Ru)

pd.DataFrame({
  'network': ['Israeli Artists', 'Sci-Fi Books', 'RuPaul Verse'],
  'n': [n_isr, n_sci, n_ru],
  'c': [c_isr, c_sci, c_ru],
  '2ln(n)/ln(c)': [2 * np.log(n_isr)/np.log(c_isr), 2 * np.log(n_sci)/np.log(c_sci), 2 * np.log(n_ru)/np.log(c_ru)],
  'd': [d_isr, d_sci, d_ru]
})
```

.insight[
`r emo::ji("bulb")` What famous phenomenon this could help explain?
]

---



## The Giant Component

.insight[
`r emo::ji("bulb")` What is the size of the largest component with $p=0$?

What is the size of the largest component with $p=1$?
]

```{python GCC-Size0, fig.show="hide"}
n = 100
s = []
for p in np.linspace(0, 1, 101):
  s_p = []
  for i in range(10):
    G = nx.erdos_renyi_graph(n = n, p = p)
    Gcc = sorted(nx.connected_components(G), key=len, reverse=True)
    Gcc_n = G.subgraph(Gcc[0]).number_of_nodes()
    s_p.append(Gcc_n / n)
  s.append(np.mean(s_p))

plt.plot(s, marker='*')
locs, labels = plt.xticks(np.linspace(0, 100, 11), np.round(np.linspace(0, 1, 11), 1))
plt.xlabel('p (where n = 100)')
plt.ylabel('GCC size S')
plt.show()
```

---

```{python GCC-Size, ref.label = "GCC-Size0", echo = FALSE, out.width = "50%"}

```

Surprisingly, for a given $n$, when $p$ is varied from 0 to 1, the giant component size $S$ suddenly "jumps"!

This is also true for a given $p$ as you increase $n$: imagine sitting at the comfort of your home watching your ER network form, when suddenly...

---

```{python GCC-Size2, echo=FALSE, out.width="50%"}
p = 0.1
s = []
for n in np.int8(np.linspace(10, 100, 100)):
  s_n = []
  for i in range(20):
    G = nx.erdos_renyi_graph(n = n, p = p)
    Gcc = sorted(nx.connected_components(G), key=len, reverse=True)
    Gcc_n = G.subgraph(Gcc[0]).number_of_nodes()
    s_n.append(Gcc_n / n)
  s.append(np.mean(s_n))

plt.plot(s, marker='*')
locs, labels = plt.xticks(np.linspace(0, 90, 10), np.int8(np.linspace(10, 100, 10)))
plt.xlabel('n (where p = 0.1)')
plt.ylabel('GCC size S')
plt.show()
```

Don't blink! Because there seems to be a *critical point*, a *transition point*, a *percolation point* where $S$ increases from 0 and finally reaches to 1 making your network fully connected.

---

### A slight wave of hands

$S$ is the probability of a node being connected to the GCC.

$S = P(\text{node i is connected to GCC})=$
$= 1-P(\text{node i is disconnected from GCC})=$
$= 1-P(\text{node i is disconnected from GCC via node j})^{n-1}$

For a given $i,j$:

$P(\text{node i is disconnected via node j})=$
$= P(\text{node i is disconnected with j}) +$
$P(\text{node i is connected with j but j is disconnected from GCC})=$
$= (1-p) + p(1-S) = 1-pS$

---

So:

$S = 1-(1-pS)^{n-1}$

And you can immediately tell finding a closed solution for $S$ would be hard.

Substituting $p=\frac{c}{n-1}$ where $c$ is the mean degree:

$S = 1-(1-\frac{c}{n-1}S)^{n-1}$

Recall that $e^x=\lim_{n\to \infty}(1+\frac {x}{n})^{n}$ so in the limit of $n$:

$S = 1-e^{-cS}$

Now there is no closed solution for $S$ but we can isolate $c=(n-1)p$:

$c=-\frac{\ln(1-S)}{S}$

---

```{python GCC-Size3, out.width="30%", echo=FALSE}
s = np.linspace(0, 1, 101)
plt.plot(s, -np.log(1-s)/s)
plt.xlabel("S", fontsize=16)
plt.ylabel("c", fontsize=16)
plt.show()
```

We can see that around $c=1$ (or $p=1/n$) the size of the GCC starts increasing .font80percent[(makes sense)].

It can be shown that around $c=\ln(n)$ (or $p=\frac{\ln(n)}{n}$) we expect the network to be fully connected.

.insight[
`r emo::ji("bulb")` So according to the ER model, for 1K nodes, a probability of 0.7% of an edge is enough to make the network connected. What do you think of that?
]

---

All of the 3 networks we've talked about are fully connected .font80percent[(though notice how the Israeli Artists and the Sci-Fi Books networks were created)].

But the model gives us a way of describing at what *stage* they are and how *robust* they are to node failures:

```{python}
pd.DataFrame({
  'network': ['Israeli Artists', 'Sci-Fi Books', 'RuPaul Verse'],
  'n': [n_isr, n_sci, n_ru],
  'c': [c_isr, c_sci, c_ru],
  'ln(n)': [np.log(n_isr), np.log(n_sci), np.log(n_ru)],
  'stage': ['subcritcial', 'supercritical', 'connected']
})
```

---

## ER Model Summary

Models well:
- Diameter, average path length
- Giant Component, Percolation, Network Robustness

Does not model well:
- Degree distribution
- Transitivity (CC)
- Communities
- Homophily

---

class: section-slide

# The Configuration Model

---

Let's see some larger real-life networks:

- The clients of a high-end Brazilian escort service, connected by whether they used the same escort
- Amazon products, scraped online, connected by whether they are recommended with each other
- Astrophysics researchers co-authorship network

```{python, echo=FALSE}
# escort = pd.read_csv('../data/escort.csv', skiprows=24,
#   sep=';', header=None, usecols=[0,1], names=['escort_id', 'client_id'])
# E = nx.from_pandas_edgelist(escort, 'client_id', 'escort_id', create_using=nx.Graph)
# E = nx.bipartite.projected_graph(E, escort['client_id'].to_list())
# df = nx.to_pandas_edgelist(E, nodelist=E.nodes)
# df.to_csv('../data/escort_clients.csv')
# isolated_clients = [node for node, degree in E.degree() if degree == 0]
# pd.DataFrame({'isolated_client_id': isolated_clients}).to_csv('../data/escort_clients_isolated.csv')

escort = pd.read_csv('../data/escort_clients.csv')
isolated_clients = pd.read_csv('../data/escort_clients_isolated.csv')['isolated_client_id'].to_list()
E = nx.from_pandas_edgelist(escort)
E.add_nodes_from(isolated_clients)

amazon = pd.read_csv("../data/com-amazon.ungraph.txt", skiprows=4, sep='\t', names=['source', 'target'])
Amz = nx.from_pandas_edgelist(amazon)

astro = pd.read_csv("../data/CA-AstroPh.txt", skiprows=4, sep='\t', names=['source', 'target'])
CA = nx.from_pandas_edgelist(astro)
```

```{python, echo=FALSE}
def size_of_GCC(G): return round(len(max(nx.connected_components(G), key=len))/G.number_of_nodes(), 2)

n_e, n_amz, n_ca = E.number_of_nodes(), Amz.number_of_nodes(), CA.number_of_nodes()
m_e, m_amz, m_ca = E.number_of_edges(), Amz.number_of_edges(), CA.number_of_edges()
p_e, p_amz, p_ca = estimate_p(m_e, n_e), estimate_p(m_amz, n_amz), estimate_p(m_ca, n_ca)
c_e, c_amz, c_ca = estimate_c(E), estimate_c(Amz), estimate_c(CA)
S_e, S_amz, S_ca = size_of_GCC(E), size_of_GCC(Amz), size_of_GCC(CA)

pd.DataFrame({
  'network': ['Escort', 'Amazon Products', 'CoAuthorship'],
  'n': [n_e, n_amz, n_ca],
  'm': [m_e, m_amz, m_ca],
  'p': [p_e, p_amz, p_ca],
  'c': [c_e, c_amz, c_ca],
  'S': [S_e, S_amz, S_ca]
})
```

---

### Degree histograms

```{python Degree-Hists2, echo=FALSE, fig.asp=0.5, out.width="80%"}
plt.figure()
plt.gcf().subplots_adjust(bottom=0.15)
plt.subplot(1, 3, 1)
degree_hist(E, xlabel='')
plt.subplot(1, 3, 2)
degree_hist(Amz, ylabel='')
plt.subplot(1, 3, 3)
degree_hist(CA, xlabel='', ylabel='')
plt.show()
```

---

class: section-slide

# Detour: The Power Law Distribution

---

### Power Law

If $X \sim PL(\alpha, x_\text{min})$, then:

$f(x) = Cx^{-\alpha}\space\mbox{for}\ x\ge x_\text{min}>0;\alpha>1$

where $C=(\alpha - 1)x^{\alpha-1}_\text{min}$

and we can write: $f(x)=\frac{\alpha-1}{x_\text{min}}(\frac{x}{x_\text{min}})^{-\alpha}$

.insight[
`r emo::ji("bulb")` What do you need to check in order to verify this is a proper density function?
]

Why do we like the Power Law distribution so much?

- scale invariance
- heavy tail
- the 80/20 (Pareto) rule

---

### Scale Invariance

If we multiply $X$ by constant $m$ the density "scales":

$f(mx)=C(mx)^{-\alpha}=Cm^{\alpha}x^{-\alpha}=C'x^{-\alpha}\propto f(x)$

Thus, $mX$ will distribute Power-Law with the same $\alpha$ parameter and a different constant.

The relative likelihood between small and large events is the same, no matter what choice of “small” we make.

.insight[
`r emo::ji("bulb")` How would $mX$ distribute if $X$ distributed Normal?
]

---

Another way of seeing this:

If we take the log-log transformation of the Power-Law distribution we get:

$\ln(f(x))=\ln(Cx^{-\alpha})=\ln(C)-\alpha\ln(x)$

Thus the log-log plot of Power-Law should show a straight line with slope $-\alpha$, and multiplying by $m$ only changes the intercept.

```{r Scale-Invariance, echo=FALSE, out.width="80%", fig.asp=0.5}
x <- seq(1, 10, 0.1)
par(mfcol = c(1, 2))
plot(x, (2.5-1)/1 * (x/1)^(-2.5), type = "l", ylab = "f(x)")
lines(x, (2.5-1)/1 * (x/1)^(-2.5)*(2^(-2.5)), col="red")
lines(x, (2.5-1)/1 * (x/1)^(-2.5)*(3^(-2.5)), col="green")
legend("topright", legend = paste0("m=", 1:3), col = 1:3, lty = 1)
plot(log(x), log((2.5-1)/1 * (x/1)^(-2.5)), type = "l", xlab = "ln(x)", ylab = "ln(f(x))")
lines(log(x), log((2.5-1)/1 * (x/1)^(-2.5)*(2^(-2.5))), col="red")
lines(log(x), log((2.5-1)/1 * (x/1)^(-2.5)*(3^(-2.5))), col="green")
legend("topright", legend = paste0("m=", 1:3), col = 1:3, lty = 1)
```

---

### Heavy Tail

The $k$-th moment of the PL distribution converges only for $k < \lfloor{\alpha-1}\rfloor$:

$E(X^k)=\int_{x_\text{min}}^{\infty}x^kf(x)=\frac{\alpha-1}{x^{\alpha-1}_\text{min}}\int_{x_\text{min}}^{\infty}x^{k-\alpha}=x^k_\text{min}(\frac{\alpha-1}{\alpha-1-k})$

So:
- For $1<\alpha\le2$ the mean is infinite (and all higher moments)
- For $2<\alpha\le3$ the mean is finite but the variance is infinite

As a result, the PL distribution is very good for modeling extreme distributions, where we *expect* extreme values, as opposed to e.g. the Normal distribution.

---

For example, here is a [post](http://giorasimchoni.com/2019/03/04/2019-03-04-on-fechner-weber-whorf-zipf-and-pareto/) of mine, showing how the PL distribution is much more suitable for modeling how words are distributed in a large corpus of text such as Wikipedia, in comparison with the Exponential distribution, both fitted via MLE:

<img src = "images/Wikipedia-Pareto-Exponential_Fit.png" style="width: 100%">

---

Another nice way of showing this is comparing $f(X)$ under the PL distribution vs. $f(X)$ under the Exponential distribution for suitable values of $\alpha$ and $\lambda$:

<img src = "images/PL-Exponential-Compare.png" style="width: 100%">

---

### 80/20 (Pareto) Rule

The Pareto rule says that "80% of the wealth is in the hands of the richest 20% of people".

But the Pareto distribution is just a different formulation of the Power Law distribution, and this phenomenon is indeed characteristic to the PL distribution.

If $X$ is "wealth" and we want to know what is the probability of being "wealthy", i.e. the part *of population* which holds at least $m$ wealth:

$P(X)=P(X>m)=1-F_m(X)=\int_{m}^{\infty}Cx^{-\alpha}dx=(\frac{m}{x_\text{min}})^{1-\alpha}$

The part *of wealth* that those "wealthy" have is the division of expectations:

$W(X) = \frac{E(X > m)}{E(X)}=\frac{\int_{m}^{\infty}xCx^{-\alpha}dx}{x^k_\text{min}(\frac{\alpha-1}{\alpha-2})}=(\frac{m}{x_\text{min}})^{2-\alpha}$

---

So, being wealthy isn't dependent on $m$, only on $\alpha$:

$W(X)=P(X)^{\frac{\alpha-2}{\alpha-1}}$

And this is how we get to the 80/20 rule without defining what wealthy is:

<img src = "images/PL-Pareto-Rule.png" style="width: 100%">

.insight[
`r emo::ji("bulb")` The dashed line shows the empirical Lorenz curve for the wealthiest individuals in the United States (data from the Forbes 400, 2003)!
]
---

class: section-slide

# End of Detour

---