---
title: "The Pandasverse"
subtitle: "Applications of Data Science"
author: "Giora Simchoni"
institute: "Stat. and OR Department, TAU"
date: "`r Sys.Date()`"
output_dir: "images"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    seal: false
    chakra: "../libs/remark-latest.min.js"
    includes:
      in_header: "../header.html"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: logo-slide

---

class: title-slide

## The Pandasverse

### Applications of Data Science - Class 6

### Giora Simchoni

#### `gsimchoni@gmail.com and add #dsapps in subject`

### Stat. and OR Department, TAU
### `r Sys.Date()`

---
```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

class: section-slide

# `Numpy`: Your best friend

---

### Python was not made for Data Science

```{python, error=TRUE, linewidth=80}
mean([1, 2, 3, 4, 5])
```

Enter `Numpy` to the rescue:

```{python}
import numpy as np

np.mean(np.array([1, 2, 3, 4, 5]))

np.array([1, 2, 3, 4, 5]).mean()
```

---

### `Numpy` Arrays

Create with a list:

```{python}
a = np.array([1, 2, 3])
print(type(a))
print(a.shape)
```

.warning[
`r emo::ji("warning")` Index is zero based!
]

```{python}
print(a[0])
```

---

Create a 2D array:

```{python}
b = np.array([[1,2,3],[4,5,6]])
print(b)
print(b.shape)
```

---

Many ways to create "typical" arrays:

```{python, eval=FALSE}
# create an array of all zeros
# (the parameter is a tuple specifying the array shape)
a = np.zeros((2,2))

# create an array of all ones
b = np.ones((1,2))

# create a constant array
c = np.full((2,2), 7)

# create a 2x2 identity matrix
d = np.eye(2)

# create an array filled with random U(0, 1) values
e = np.random.random((2,2))

# create a sequence from 2 to 15, not including
np.arange(2, 15)

# create sequence of 11 numbers between 0 and 1 including
np.linspace(0, 1, 11)
```

---

And every array has a `reshape()` method:

```{python}
np.arange(0.1, 1, step=0.1).reshape(3, 3)
```

---

### `Numpy` Math

Elementwise multiplication:

```{python}
x = np.array([[1,2],[3,4]], dtype=np.float64)
y = np.array([[5,6],[7,8]], dtype=np.float64)
print(x * 2)
```

Elementwise sum:
```{python}
print(x + y)
```

Same:

```{python, eval=FALSE}
print(np.add(x, y))
```

---

You get the idea:

```{python, eval=FALSE}
print(x - y)
print(np.subtract(x, y))

print(x * y)
print(np.multiply(x, y))

print(x / y)
print(np.divide(x, y))

print(np.sqrt(x))
```

---

Vector/Matrix multiplication:

```{python}
print(x.dot(y))
print(np.dot(x, y))

v = np.array([9,10])
w = np.array([11, 12])

print(v.dot(w))
print(np.dot(v, w))
```

---

Transpose

```{python}
x = np.array([[1,2],[3,4]])
print(x.T)
```

Sum, mean, std, median, quantile, min, max...:

```{python}
print(np.sum(x))  # Compute sum of all elements
print(np.sum(x, axis=0))  # Compute sum of each column
print(np.std([1,2,3])) # possible, in case you were wondering
```

---

### `Numpy` Indexing and Slicing

Similar to R but there are some things worth noticing:

```{python}
a = np.array([[1,2,3,4], [5,6,7,8], [9,10,11,12]])
print(a)
```

```{python}
# use slicing to pull out the subarray consisting of the first 2 rows
# and columns 1 and 2; b of shape (2, 2)
b = a[:2, 1:3]
print(b)
```

---

```{python}
# a slice of an array is a view into the same data, so modifying it
# will modify the original array.
print(a[0, 1])
b[0, 0] = 77
print(a[0, 1])
```

---

Very convenient, R does not have these features without external packages:

```{python}
# index "from last place"
a[-2:]
```

```{python}
# reverse an array
a = np.arange(5)
print(a[::-1])
```

---

Working with boolean masks like in R:

```{python}
print(a[a > 2])
print(a[np.where(a > 2)])
print(a[np.argmin(a)])
```

---

class: section-slide

# `Scipy`: Scientific Computing and Stats

---

### Many modules, let's focus on:

* **`sparse`**: Sparse Matrices manipulation
* **`ndimage`**: Images manipulation (though see `scikit-image` and `opencv`)
* **`stats`**: Statistics (though see `statsmodels`)

---

### `sparse`

```{python}
from scipy.sparse import csr_matrix

row = np.array([0, 0, 1, 2, 2, 2])
col = np.array([0, 2, 2, 0, 1, 2])
data = np.array([1, 2, 3, 4, 5, 6])
sparse_a = csr_matrix((data, (row, col)), shape=(3, 3))

print(sparse_a.toarray())
```

---

### `ndimage`

```{python}
from scipy import ndimage
from scipy import misc
import matplotlib.pyplot as plt

face = misc.face(gray=True)
blurred_face = ndimage.gaussian_filter(face, sigma=10)

print(face.shape)
print(face[:5, :5])
```

---

```{python Face, out.width="100%", fig.asp=0.5}
plt.subplot(121)
plt.imshow(face, cmap=plt.cm.gray)
plt.subplot(122)
plt.imshow(blurred_face, cmap=plt.cm.gray)
plt.show()
```

---

### `stats`

```{python}
from scipy import stats

rvs1 = stats.norm.rvs(loc=5,scale=10,size=500)
rvs2 = stats.norm.rvs(loc=5,scale=10,size=500)

stats.ttest_ind(rvs1,rvs2)
```

---

class: section-slide

# `Pandas`: Data, Data, Data

---

### Pandas: not `data.frame`, `DataFrame`

```{r, results="asis", echo=FALSE}
tab <- "| `dplyr`	| `pandas` |
|------------------|
|`mutate` | `assign`|
|`select` | `filter`|
|`rename` | `rename`|
|`filter` | `query`|
|`arrange` | `sort_values`|
|`group_by` | `groupby`|
|`summarize` | `agg`|"

cat(tab)
```

.insight[
`r emo::ji("bulb")` Thare *are* Pandas dialects, don't go translating your pipes verbatim.
]

---

### Read in the data

```{python}
import pandas as pd

okcupid = pd.read_csv("../data/okcupid.csv.zip")
```

```{python}
okcupid.shape
okcupid.columns
```

---

### `info()` and `describe()`

```{python}
okcupid.describe()
```

---

### `assign()`

Add a column `height_cm`, the `height` in centimeters:

```{python}
okcupid = okcupid.assign(height_cm = okcupid['height'] * 2.54)

okcupid = okcupid.assign(height_cm = lambda x: x.height * 2.54)
```

If you don't need a pipe just do:

```{python}
okcupid['height_cm'] = okcupid['height'] * 2.54
```

---

## `query()` and `filter()`

Query only women, filter only age and height:

```{python}
okcupid \
  .query('sex == "f"') \
  .filter(['age', 'height']) \
  .head(5)
```

Again, without a pipe:

```{python, eval=FALSE}
okcupid[okcupid['sex'] == "f"][['age', 'height']]
```

---

Same but income over 100K, and select all essay questions:

```{python}
okcupid \
  .query('sex == "f" and income > 100000') \
  .filter(okcupid.columns[okcupid.columns.str.startswith('essay')])
```

---

### `agg()`

Find the average height of women

```{python}
okcupid \
  .query('sex == "f"') \
  .filter(['height_cm']) \
  .agg('mean')
```

Notice we got a `pd.Series`, the `Pandas` equivelent for a vector. We could use the `.values` attribute to pull the `Numpy` array behind the Series:

```{python}
okcupid \
  .query('sex == "f"') \
  .filter(['height_cm']) \
  .agg('mean').values
```

---

### `groupby()`

But why settle for women only?

```{python}
okcupid \
  .groupby('sex')['height_cm'] \
  .agg('mean')
```

And you might want to consider `rename()`ing sex!

```{python}
okcupid \
  .groupby('sex')['height_cm'] \
  .agg('mean') \
  .rename_axis(index = {'sex': 'gender'})
```

---

Group by multiple variables, get more summaries, arrange by descending average height:

```{python}
okcupid \
  .groupby(['sex', 'status'])['height_cm'] \
  .agg(['mean', 'median', 'count']) \
  .sort_values('median', ascending=False)
```

---

### Pro tip: `size()`

When all you want is, well, `size`:

```{python}
okcupid.groupby('body_type').size()
```

---

### `loc`, `iloc` and `at`

---

class: section-slide

# `seaborn`: Visualization

---

```{python Seaborn-Scatterplot}
import matplotlib.pyplot as plt
import seaborn as sns

sns.set()
g = sns.scatterplot('age', 'height_cm', hue='sex', data = okcupid)
plt.show()
```

---

```{python Seaborn-Relplot, out.width="100%"}
g = sns.relplot('age', 'height_cm',
  hue = 'sex', kind = 'scatter', col='sex', data = okcupid)
plt.show()
```

