---
title: "Wrangling Data in the Tidyverse"
subtitle: "Applications of Data Science"
author: "Giora Simchoni"
institute: "Stat. and OR Department, TAU"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    seal: false
    includes:
      in_header: "../header.html"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: logo-slide

---

class: title-slide

## Wrangling Data in the Tidyverse

## Applications of Data Science - Class 1

### Giora Simchoni

#### `gsimchoni@gmail.com and add #dsapps in subject`

### Stat. and OR Department, TAU
### `r Sys.Date()`

---
```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

# So, what's wrong with Excel?

(MS Excel is one amazing software. But it has no:)

- Structure (or rather, structure is up to the user)
- Types to variables
- Automization (you could learn VBA Excel, but the horror)
- Reproducibility
- Open Source
- Extensibility
- Speed and Scale
- Modeling (there *is* a t-test, but the horror)

[MS Excel might be the most dangerous software on the planet](https://www.forbes.com/sites/timworstall/2013/02/13/microsofts-excel-might-be-the-most-dangerous-software-on-the-planet/#667084f5633d) (Tim Worstall, Forbes)

---

# So, what's wrong with base R?