---
title: "Topics in Classification"
subtitle: "Applications of Data Science"
author: "Giora Simchoni"
institute: "Stat. and OR Department, TAU"
date: "`r Sys.Date()`"
output_dir: "images"
output:
  xaringan::moon_reader:
    css: "../slides.css"
    seal: false
    chakra: "../libs/remark-latest.min.js"
    includes:
      in_header: "../header.html"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: logo-slide

---

class: title-slide

## Topics in Classification

### Applications of Data Science - Class 13

### Giora Simchoni

#### `gsimchoni@gmail.com and add #dsapps in subject`

### Stat. and OR Department, TAU
### `r Sys.Date()`

---
```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

class: section-slide

# Topics in Classification

---

# Life isn't perfect

Let's tackle just a few issues:

- Imbalanced Classes
- Multiple Classes
- Not enough labelled data and data labeling is expensive

---

class: section-slide

# Imbalanced Classes

---

### Typical examples of Imbalanced Classes scenarios

- Rare diseases: [this](https://www.kaggle.com/c/hivprogression) dataset contains genetic data for 1,000 HIV patients, 206 out of 1,000 patients improved after 16 weeks of therapy
- Conversion/Sell/CTR rates: [this]() dataset contains 10 days of Click-Through-Rate data for Avazu mobile ads, ~6.8M clicked out of ~40.4M
- Fraud detection: [this](https://www.kaggle.com/mlg-ulb/creditcardfraud) dataset contains credit card transactions for a major European CC, 492 frauds out of 284,807 transactions
