---
title: 'NetworkX'
subtitle: 'Applications of Data Science'
author: 'Giora Simchoni'
institute: 'Stat. and OR Department, TAU'
date: '`r Sys.Date()`'
output_dir: 'images'
output:
  xaringan::moon_reader:
    css: '../slides.css'
    seal: false
    chakra: '../libs/remark-latest.min.js'
    includes:
      in_header: '../header.html'
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: logo-slide

---

class: title-slide

## NetworkX

### Applications of Data Science - Class 7

### Giora Simchoni

#### `gsimchoni@gmail.com and add #dsapps in subject`

### Stat. and OR Department, TAU
### `r Sys.Date()`

---
```{r child = '../setup.Rmd'}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

printmrow <- function(x) cat(cat(x,sep=' & '),'\\\\')

matrix2latex <- function(matr) {
  cat('$\\begin{bmatrix}')
  body <- apply(matr,1,printmrow)
  cat('\\end{bmatrix}$')
}
```

class: section-slide

# Why NetworkX?

---

## Pros and Cons of `NetworkX`

Pros:
- Maintained!
- Friendly: Community, Docs, Installation (as opposed to `igraph`)
- Relatively easy interface (as opposed to `igraph`)
- Speaks pandas, numpy and other important python tools
- Nodes can be anything, including python objects

Cons:
- Slow, probably not for large networks (as opposed to `igraph`, `graph-tool`)
- Unimpressive viz (as opposed to Gephi, ggraph, others)
- Not the most comprehensive algorithms selection

---

class: section-slide

# Creating a Graph(): Manually

---

## Four types of Graph objects

Undirected:

```{python}
import networkx as nx

G = nx.Graph()
```

Directed:

```{python}
G = nx.DiGraph()
```

Multiedge undirected:

```{python}
G = nx.MultiGraph()
```

Multiedge directed:

```{python}
G = nx.MultiDiGraph()
```

---

## Add Nodes

One or more nodes:

```{python}
G = nx.Graph()
G.add_node('John')
G.add_nodes_from(['Paul', 'George', 'Ringo'])
```

Watch a Graph's nodes:
```{python}
G.nodes
```

Print nicer as a list:

```{python}
list(G.nodes)
```

---

## Add Edges

One or more edges:

```{python}
G.add_edge('John', 'Paul')
G.add_edges_from([('Paul', 'George'), ('John', 'Ringo')])
```

Surprisingly, this would also work:

```{python}
G.add_edge('John', 'Brian')
```

Watch a Graph's edges:
```{python}
print(list(G.edges))
```

---

## Remove nodes or edges

```{python}
G.remove_node('Brian')
```

Did `NetworkX` bother to remove the edge as well?

```{python}
print(list(G.edges))
```

Remove an edge, remove from a list:

```{python}
G.remove_edge('Paul', 'John') # why did it work?
G.remove_nodes_from(['John', 'George'])
```

Break the Beatles altogether:

```{python}
G.clear()
```

---

## Good to know

- Adding an edge also adds its nodes if they didn't exist before
- Adding a duplicate node/edge is ignored unless `MultiGraph()`
- Removing an edge does not remove its nodes
- Removing a node removes all edges to/from it
- Removing a non-existent node/edge raises an error unless it is part of a list in which case - ignored

---

## Nodes and Edges Attributes

While adding to Graph:

```{python}
G.add_node('Ringo', alive=True)
G.add_nodes_from([('George', {'alive': False}), ('John', {'alive': False})])
# can also do: G.add_nodes_from(['George', 'John'], alive=False)
G.add_edge('John', 'Paul', year=1957)
G.add_edges_from([('Ringo', 'John'), ('Ringo', 'Paul'), ('Ringo', 'George')], year=1960)
```

Watch attributes with the `nodes()` and `edges()` methods:

```{python}
print(G.nodes(data=True))
print(G.nodes(data='alive'))
```

---

## Nodes and Edges Attributes

Setting an attribute of an existing node/edge through the `nodes` and `edges` dictionary attributes:

```{python}
print(G.nodes['Paul'])
G.nodes['Paul']['alive'] = True #?
```

```{python}
G.add_edge('John', 'George')
G.edges[('John', 'George')]['year'] = 1958
```

As a dictionary you can also delete an attribute this way:

```{python}
del G.nodes['Paul']['alive']
del G.edges[('John', 'George')]['year']
```

---

## Nodes and Edges Attributes

Finally you can set an attribute from a simple dictionary:

```{python}
year_met = {
  ('John', 'Paul'): 1957,
  ('Paul', 'Ringo'): 1960,
  ('Paul', 'George'): 1958,
  ('John', 'George'): 1958
}
nx.set_edge_attributes(G, year_met, 'year')
```

Or multiple attributes with a nested dictionary:

```{python}
node_attrs = {
  'John': {'alive': False, 'born': 1940},
  'Paul': {'alive': True, 'born': 1942},
  'George': {'alive': False, 'born': 1943},
  'Ringo': {'alive': True, 'born': 1940}
}
nx.set_node_attributes(G, node_attrs)
```

---

### Good to have

The edge attribute `weight` is so important it got its own method:

```{python}
G = nx.Graph()
G.add_weighted_edges_from([(0, 1, 3.0), (1, 2, 7.5)])
```

Quick methods to know:

```{python}
print(G.number_of_nodes())
print(G.number_of_edges())
print(G.is_directed())
```

---

### Good to have

```{python}
print(G.has_node(3))
print(G.has_edge(0, 1))
print(G.subgraph([0, 1]).number_of_edges())
print(list(G.neighbors(1)))
```

---

## What about `DiGraph()`?

Convert an undirected graph to a directed graph:

```{python}
D = nx.DiGraph(G)
```

Edges direction matters:

```{python}
D.add_edge(2, 3, weight=10.0)
print(D.has_edge(2, 3))
print(D.has_edge(3, 2))
```

Other than that...

---

class: section-slide

# Visualize a Graph(): matplotlib

---

### `draw()`: The very basic

```{python, Basic, out.width="45%"}
import matplotlib.pyplot as plt

G = nx.erdos_renyi_graph(n=5, p=0.6) # a.k.a binomial_graph()

nx.draw(G)
plt.show()
```

---

### `draw_networkx()`: Some more options

```{python, Basic2, out.width="45%"}
nx.draw_networkx(G, node_size=800, node_color='red', edge_color='pink',
  width=5, font_size=16)
plt.show()
```

---

You may like the unpacking a dictionary option better:

```{python, Basic3, out.width="45%"}
options = {
    'node_color': 'black',
    'node_size': 100,
    'width': 3,
}

nx.draw_networkx(G, **options)
plt.show()
```

---

### Layouts Galore

```{python, Basic4, out.width="45%"}
plt.subplot(121)
limits = plt.axis('off')
nx.draw_networkx(G, pos=nx.kamada_kawai_layout(G))
plt.subplot(122)
limits = plt.axis('off')
nx.draw_networkx(G, pos=nx.fruchterman_reingold_layout(G))
plt.show()
```

---

### Use a Layout directly

```{python, Basic5, out.width="45%"}
nx.draw_kamada_kawai(G)
plt.show()
```

---

### For nicer graphs go to...

- [Grave](https://networkx.github.io/grave/)?
- [nxviz](https://nxviz.readthedocs.io/en/latest/index.html)?
- [Plotly](https://plot.ly/python/) (interactive graphs for Python or R)
- [Gephi](https://gephi.org/) (a Desktop app, neither R nor Python but both export to Gephi format)
- [d3](https://d3js.org/) (JS delight)
- [ggraph](https://ggraph.data-imaginist.com/) (R)?
---

class: section-slide

# Create a Graph(): In Real Life

---

### Numpy

```{python Numpy, out.width="30%"}
import numpy as np

A = np.array([
[0, 1, 1, 0, 0],
[1, 0, 1, 0, 1],
[1, 1, 0, 0, 1],
[0, 0, 0, 0, 0],
[0, 1, 1, 0, 0]])

G = nx.from_numpy_matrix(A)
nx.draw_networkx(G, node_size=0, font_size=40)
plt.show()
```

---

### Edgelist - a list of edges

```{python Beatles1, out.width="40%"}
beatles_edgelist = [('John', 'Paul', {'weight': 1.0}), ('John', 'George', {'weight': 0.5}),
  ('John', 'Ringo', {'weight': 0.5})]

G = nx.from_edgelist(beatles_edgelist)

weights = [10 * attrs['weight'] for u, v, attrs in G.edges(data=True)]
nx.draw_networkx(G, width=weights)
plt.show()
```

---

### Edgelist - a file

This is how the `beatles.edgelist` file looks like:

```{python, eval=FALSE}
John Paul 1
John George 0.5
John Ringo 0.5
```

Read it with `read_edgelist()` or `read_weighted_edgelist()`:

```{python}
with open('../data/beatles.edgelist', 'rb') as edgelist_file:
  G = nx.read_edgelist(edgelist_file, data=(('weight', float),))
print(G.edges(data=True))
```

```{python}
with open('../data/beatles.edgelist', 'rb') as edgelist_file:
  G = nx.read_weighted_edgelist(edgelist_file)
print(G.edges(data=True))
```

---

### CSV

This is how the `beatles.csv` file looks like:

```{python, eval=FALSE}
John,Paul,1957,1.0
John,George,1958,0.5
John,Ringo,1960,0.5
```

You can still use `read_edgelist()` to read it:

```{python}
with open('../data/beatles.csv', 'rb') as edgelist_file:
  G = nx.read_edgelist(edgelist_file, delimiter=',', data=(('year', int),('weight', float),))
print(G.edges(data=True))
```

Or, you can just use Pandas `read_csv()` and...

---

### Pandas

```{python Beatles2, out.width="40%"}
import pandas as pd

beatles_df = pd.read_csv('../data/beatles.csv', header=None)
beatles_df.columns = ['source', 'target', 'year', 'weight']

G = nx.from_pandas_edgelist(beatles_df, 'source', 'target', ['year', 'weight'])

nx.draw_networkx(G)
plt.show()
```

---

### What about writing?

All methods we've seen have a `write_` or `to_` complementary function:

```{python}
print(nx.to_edgelist(G))
```

### What about `DiGraph()`?

All methods we've seen have a `create_using` parameter:

```{python}
D = nx.from_pandas_edgelist(beatles_df, 'source', 'target',
  ['year', 'weight'], create_using=nx.DiGraph)
```

.warning[
`r emo::ji("warning")` When creating `DiGraph()` NetworkX assumes $A_{ij}$ corresponds to the edge from i to j, contrary to our convention. Therefore, you should use `A.transpose()`
]
---

### Bipartite

For creating a bipartite network, use the `bipartite` node attribute:

```{python Bipartite1, out.width="40%"}
G = nx.Graph()
G.add_nodes_from([1, 2, 3, 4], bipartite=0)
G.add_nodes_from(['a', 'b', 'c'], bipartite=1)
G.add_edges_from([(1, 'a'), (1, 'b'), (2, 'b'), (2, 'c'), (3, 'c'), (4, 'a')])

nx.draw_networkx(G, pos=nx.bipartite_layout(G, [1, 2, 3, 4]), node_size=0, font_size=20)
plt.show()
```

It is your responsibility though to check that indeed the two sets of nodes are not connected.

---

"In real life" you would probably have something like the Marvel incidence matrix in a CSV, what `NetworkX` calls a *biadjacency_matrix*, or a list of edges you could make a Scipy *sparse matrix* with, from which you can create a bipartite graph:

```{python Bipartite2}
from scipy import sparse

marvel = pd.read_csv("../data/marvel_incidence_matrix.csv")
B = marvel.iloc[:, 1:].values

sB = sparse.csr_matrix(B)

G = nx.bipartite.from_biadjacency_matrix(sB)
print(G.nodes(data=True))
```

---

```{python Bipartite3, out.width = "50%"}
films = marvel['Film'].to_list()
characters = marvel.columns[1:].to_list()

id_to_label = {id: char for id, char in enumerate(films + characters)}
G = nx.relabel_nodes(G, id_to_label)

nx.draw_networkx(G, pos=nx.bipartite_layout(G, films), node_size=0, font_size=10)
plt.show()
```

---

You've seen how to project an incidence matrix with numpy:

```{python Bipartite4, out.width = "50%"}
P = B.transpose() @ B
np.fill_diagonal(P, 0)
Gc = nx.from_numpy_matrix(P)

id_to_char = {id: char for id, char in enumerate(characters)}
Gc = nx.relabel_nodes(Gc, id_to_char)
nx.draw_networkx(Gc, node_size=0, font_size=10)
plt.show()
```

---

But NetworkX can do all that for you:

```{python Bipartite5, out.width = "50%"}
Gc = nx.bipartite.projected_graph(G, characters)

nx.draw_networkx(Gc, node_size=0, font_size=10)
plt.show()
```

---

### Similarity

Let's recreate the sci-fi similarity network. We'll do some filtering:
- The original CSV includes 156 books, we'll take only those marked as "popular"
- Take only books with at least 2 features marked > 0
- Won't be looking at the author's name, gender, or date of release

```{python}
scifi = pd.read_csv('../data/sci_fi_books.csv', index_col='book')
scifi['n_features'] = (scifi.iloc[:, 6:17] > 0).sum(axis=1)
scifi = scifi[scifi['popular'] == 1]

scifi_fil = scifi[scifi['n_features'] >= 2]
cols_to_drop = ['date', 'author', 'frequency', 'author_gender',
  'quarter_century', 'century', 'popular', 'n_features']
scifi_fil = scifi_fil.drop(cols_to_drop, axis=1)

# to use Pandas corr function need to transpose data frame
scifi_corr = scifi_fil.T.corr(method='spearman')
```

---

We reached a 25x25 symmetric correlation matrix for 25 books:

```{python}
scifi_corr.values[:5, :5]
```

Turn all upper traingle to `Nan`, so we'll get only 25 * 24 / 2 correlations:

```{python}
scifi_corr.values[np.triu_indices_from(scifi_corr.values)] = np.nan

scifi_corr.values[:5, :5]
```

---

Convert this to a long edgelist, filter out the `Nan`:

```{python}
scifi_edgelist = scifi_corr.reset_index().melt(id_vars='book',
  value_vars=scifi_fil.index, var_name='book2', value_name='corr')

scifi_edgelist = scifi_edgelist[~pd.isna(scifi_edgelist['corr'])]
print(scifi_edgelist.shape)

scifi_edgelist.head(5)
```

---

Filter all correlations above 0.5 and convert to a `Graph()`:

```{python Sci-Fi, out.width="45%"}
G = nx.from_pandas_edgelist(scifi_edgelist[scifi_edgelist['corr'] > 0.5],
  'book', 'book2', ['corr'])

nx.draw_networkx(G)
limits = plt.axis('off')
plt.show()
```

---

class: section-slide

# Let NetworkX work

---

### Adjacency Matrix

You can get a Numpy matrix with `to_numpy_matrix()`:

```{python}
A = nx.to_numpy_matrix(G)

print(A.shape)

A[:4, :4]
```

.insight[
`r emo::ji("bulb")` Say, didn't we have 25 books?
]

---

But a realistic network would be large and sparse so the `adjacency_matrix()` method returns a SciPy sparse matrix:

```{python}
A = nx.adjacency_matrix(G)

print(A.shape)

print(A[:4, :4])
```

.insight[
`r emo::ji("bulb")` Do you know what a sparse matrix is?
]

---

### Easy to check if...

```{python}
print(nx.is_directed(G))
print(nx.is_directed_acyclic_graph(G))
print(nx.is_bipartite(G))
print(nx.is_connected(G))
```

---

### Degree and Density

Degree is a big deal, there's an attribute *and* a method for that:

```{python}
G.degree
```

```{python}
list(G.degree)
```

Average degree:

```{python}
np.mean(list(dict(G.degree).values()))
```

---

Get the degree of a specific node:

```{python}
G.degree['Contact']
```

Get the density of a network:

```{python}
nx.density(G)
```

For directed graphs, use `in_degree` and `out_degree`:

```{python, eval=FALSE}
D = nx.DiGraph(G)

D.in_degree()
D.out_degree()
```

---

### Paths and Diameter

Is there a path between two nodes?

```{python}
nx.has_path(G, 'Frankenstein', 'Contact')
```

What is the shortest path distance between two nodes:

```{python}
nx.shortest_path_length(G, 'Frankenstein', 'Contact')
```

Give me (one, there could be more) shortest path:

```{python}
nx.shortest_path(G, 'Frankenstein', 'Contact')
```

---

Give me all shortest paths between two nodes:

```{python}
all_sps = nx.all_shortest_paths(G, 'Frankenstein', 'Contact') # generator!
list(all_sps)
```

Get the diameter of a network:

```{python}
nx.diameter(G)
```

---

### Components

How many connected components?

```{python}
nx.number_connected_components(G)
```

All connected components, each is a set of nodes:

```{python}
cc = nx.connected_components(G) # generator!
list(cc)
```

Get component of a specific node:

```{python}
nx.node_connected_component(G, 'Contact')
```

---

Get largest component (a.k.a Giant Connected Component)

```{python}
gcc = max(nx.connected_components(G), key=len)
len(gcc)
```

For directed graphs use `strongly_`/`weakly_` prefixes with previous functions:

```{python, eval=FALSE}
D = nx.DiGraph(G)

nx.number_strongly_connected_components(D)
nx.strongly_connected_components(D)
```

---

### Laplacian

```{python}
L = nx.laplacian_matrix(G)

L.todense()[:5, :5]
```

